Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-02-05 response.tex

\maketitle

\section*{Reviewer 1}\label{sec:reviewer-1}

\begin{enumerate}
\item\textbf{Comment:} \label{comment:shortersec2}
	This paper is concerned with carrier phase and amplitude estimation for PSK modulated signalling.
The considered least squares estimator seems to be only a minor extension of the one from [7]. Yet, more than 4 pages are dedicated to its derivation. Perhaps this section could be made more concise.
	\\\textbf{Response:}
%\comment{look at shortening sec 2}

The description of the algorithm has been left intact.  An important reason for describing the algorithm is that much of the notation used during the analysis of the estimator is most naturally introduced during the description of the algorithm.  For example, the proof of Lemma~\zref{lem:fracpartlambdahatnotpi} makes use of the notation $e_i$ and the function $LS(\{d_i, i \in D\})$ introduced during the description of the algorithm.  To qualify this for the reader we have included the following sentences at the beginning of Section~\zref{sec:least-squar-estim}:

@@

``In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that will be used for analysing the estimator in Section~\zref{sec:stat-prop-least} is most naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.''

%	We have shortened Sec.\ II to make it more concise and improve readability.
\item\label{R1it2}\textbf{Comment:} 
	As stated by the authors themselves, the key results and main contributions of the paper involve the asymptotic properties of the least squares estimator. As far as I can see the derivation seems correct and complete, yet it is mathematically very involved and rather difficult to follow.
	\\\textbf{Response:}	
	For improved readability of the manuscript, we have moved all of the lemmas into the appendix. We have also added a discussion of the implications and the practical relevance of the theorems after the statement of Theorems~\zref{thm:consistency} and~\zref{thm:normality} in Section~\zref{sec:stat-prop-least}.  The introduction has been expanded to make the practical and theoretical importance of our of results clear.

@@

``This appears to be the first time that the statistical properties of a non-data-aided estimator for carrier amplitude have been analysed theoretically.''

and in the new introduction we affirm that

``while the specific results given here apply only to the least square estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.''

Thus, the techniques in our paper can potentially be deployed to accurately model the bias and variance of a non-data-aided SNR estimator in order to implement the reviewers proposal.  We confirm, however, that analysing and implementing the reviewers suggested technique is beyond the scope of the present paper which places primary interest in analysing the least squares estimator.

### 2014-02-11 paper.tex
minor
added Andre's comments

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$.  By definition $\rho_0$ is a positive real number.  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990,580211,974597,Mengali_andre_synchro_book}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:LSdefn}
\begin{split}
LS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

@@

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered what is called \emph{non-data-aided detection} (also know as \emph{blind}, \emph{noncoherent} or \emph{non-decision-directed detection}) where no pilot symbols exist ($P = \emptyset$).  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the non-data-aided case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:LSdefn}.  In the literature, analytic study of non-data-aided estimators of carrier phase has primarily involved the derivation of Cram\'{e}r-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cram\'{e}r-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors knowledge, all of the existing bounds are applicable only in the fully non-data-aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

@@

That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing systems parameters, such as the number and placement of pilot symbols, and also provides useful information to other reciever components, such as channel tracking systems.

@@

In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non-data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that will be used for analysing the estimator in Section~\ref{sec:stat-prop-least} is most naturally introduced during the description of the algorithm.  For this reason, and so that the paper is self contained, we describe the algorithm in full.  For the purpose of analysing computational complexity it is assumed that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

@@

Figures~\ref{fig:plotphaseBPSK} and~\ref{fig:plotphaseQPSK} display the non-data aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996}.  To the authors knowledge, simple expression for the CRB for $8$-PSK do not exist in the literature and so the CRB is only shown when $M=2$ (BPSK) and $M=4$ (QPSK).  These CRBs are derived for the non-data aided setting and are only valid for estimators that do not make use of pilots symbols, i.e., $\abs{P}=\emptyset$.

@@

\begin{lemma}~\label{lem:diffatlambdaL}
The derivative of $G_L$ exists, and is equal to zero, at $\hat{\lambda}_L$.  That is,
\[
G_L^\prime(\hat{\lambda}_L) = \frac{d G_L}{d \lambda}(\hat{\lambda}_L) = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that 
\[
G_L(\lambda) = \frac{1}{L}\sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}
\] 
is differentiable everywhere except when $\fracpart{\lambda + \Phi_i} = -\tfrac{\pi}{M}$ for any $i \in D$ with $R_i > 0$.  Let $q_i$ be the smallest number from the interval $[-\tfrac{\pi}{M}, 0]$ such that
\[
 \sin(q_i) > -\frac{\rho_0 \sabs{\eta}^2 R_i}{4 L \hat{\rho} \sin(\pi/M)}
\]
where $\eta = e^{-j2\pi/M} - 1$.  Observe that $q_i < 0$ when $R_i > 0$.  Lemma~\ref{lem:fracpartlambdahatnotpi} shows that
\[
\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{\pi}{M} + q_i
\]
for all $i \in D$.  Thus, $\sfracpart{\hat{\lambda}_L + \Phi_i} \neq -\tfrac{\pi}{M}$ for $i \in D$ such that $R_i > 0$ and therefore $G_L$ is differentiable at $\hat{\lambda}_L$.  That $G_L^\prime(\hat{\lambda}_L) = 0$ follows since $\hat{\lambda}_L$ is a maximiser of $G_L$.
\end{IEEEproof}

>>>>

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$.  By definition $\rho_0$ is a positive real number.  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990,580211,974597,Mengali_andre_synchro_book}.  A practical approach is the least squares estimator, that is, the minimiser of the sum of squares function
\begin{equation}\label{eq:LSdefn}
\begin{split}
LS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

@@

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered what is called \emph{non-data-aided detection} (also know as \emph{blind}, or \emph{non-decision-directed detection}) where no pilot symbols exist ($P = \emptyset$).  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the non-data-aided case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:LSdefn}.  In the literature, analytic study of non-data-aided estimators of carrier phase has primarily involved the derivation of Cram\'{e}r-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cram\'{e}r-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

@@

That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing system parameters, such as the number and placement of pilot symbols, and also provides useful information to other reciever components, such as channel tracking systems.

@@

In this section we derive Mackenthun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non-data-aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that will be used for analysing the estimator in Section~\ref{sec:stat-prop-least} is most naturally introduced during the description of the algorithm.  For this reason, and so that the paper is self-contained, we describe the algorithm in full.  For the purpose of analysing computational complexity it is assumed that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

@@

Figures~\ref{fig:plotphaseBPSK} and~\ref{fig:plotphaseQPSK} display the non-data aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996}.  To the authors' knowledge, simple expression for the CRB for $8$-PSK do not exist in the literature and so the CRB is only shown when $M=2$ (BPSK) and $M=4$ (QPSK).  These CRBs are derived for the non-data aided setting and are only valid for estimators that do not make use of pilots symbols, i.e., $\abs{P}=0$.

@@

\begin{lemma}~\label{lem:diffatlambdaL}
The derivative of $G_L$ exists, and is equal to zero, at $\hat{\lambda}_L$.  That is,
\[
G_L^\prime(\hat{\lambda}_L) = \frac{d G_L}{d \lambda}(\hat{\lambda}_L) = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that 
\[
G_L(\lambda) = \frac{1}{L}\sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}
\] 
is differentiable everywhere except when $\fracpart{\lambda + \Phi_i} = -\tfrac{\pi}{M}$ for any $i \in D$ with $R_i > 0$.  Let $q_i \in [-\tfrac{\pi}{M}, 0]$ satisfy
\[
 \sin(q_i) = -\frac{\rho_0 \sabs{\eta}^2 R_i}{4 L \hat{\rho} \sin(\pi/M)}
\]
where $\eta = e^{-j2\pi/M} - 1$.  Observe that $q_i < 0$ when $R_i > 0$.  Lemma~\ref{lem:fracpartlambdahatnotpi} shows that
\[
\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{\pi}{M} + q_i
\]
for all $i \in D$.  Thus, $\sfracpart{\hat{\lambda}_L + \Phi_i} \neq -\tfrac{\pi}{M}$ for $i \in D$ such that $R_i > 0$ and therefore $G_L$ is differentiable at $\hat{\lambda}_L$.  That $G_L^\prime(\hat{\lambda}_L) = 0$ follows since $\hat{\lambda}_L$ is a maximiser of $G_L$.
\end{IEEEproof}
